Generating Random Logic Programs Using Constraint Programming

### 2019-12-29 paper/paper.tex
Preliminary attempt at counting the number of possible programs

\section{Atoms}

\begin{definition} \label{def:atom}
  An \emph{atom} is a $\variable{predicate} \in \tokens{} \cup \predicates{}$ and a
  list of \variable{arguments} of length $\maxArity{}$ in $\variables{} \cup
  \constants{} \cup \{ \Box \}$, where $\Box$ means the position is either
  reserved for a variable, or disabled. The atom's \variable{arity} is a number
  in $[0, \maxArity{}]$ defined by a \variable{table} constraint,
  according to the \variable{predicate}.
\end{definition}

@@

>>>>

\section{Atoms} % TODO: rename to 'nodes'

%, where $\Box$ means the position is either
%  reserved for a variable, or disabled

\begin{definition} \label{def:atom}
  An \emph{atom} is a $\variable{predicate} \in \tokens{} \cup \predicates{}$
  and a list of \variable{arguments} of length $\maxArity{}$ in $\variables{}
  \cup \constants{}$. The atom's \variable{arity} is a number in $[0,
  \maxArity{}]$ defined by a \variable{table} constraint, according to the
  \variable{predicate}.
\end{definition}

@@

\section{Counting Programs}

Let $p_a$ be the number of predicates in $\predicates{}$ with arity $a \in
\arities{}$.

Number of atoms:
\[
  A = \sum_{a \in \arities{}} p_a(|\variables{}| + |\constants{}|)^a
\]

Number of clauses:
\[
  C = 1 + \sum_{n=1}^{\maxNumNodes{}} T(n),
\]
where $T(n)$ is defined recursively as:
\[
  T(1) = A
\]
and
\[
  T(n) = T(n-1) + 2\sum_{\substack{\text{ordered partitions}\\
      c_1 + \dots + c_k = n - 1,\\
      k \ge 2}} \prod_{i=1}^k T(c_i).
\]

Example of ordered partitions:
\[
  3 = 2 + 1 = 2 + 1 = 1 + 1 + 1,
\]
so for $n=4$, the sum would have three terms.

Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v=1}^a \binom{a}{v} |\constants{}|^{a-v}
  \sum_{k=0}^{|\variables{}|-1} \binom{v-1}{k}.
\]
First, select the $v$ positions dedicated for variables. The remaining $a-v$
constants can then be filled in $|\constants{}|^{a-v}$ ways. Filling $v$
positions with $k+1$ variables in a non-decreasing manner (without skipping any
variables) can be seen as putting $k$ `bars' in the $v-1$ spaces between $v$
positions. Each bar represents switching to the next variable.
% TODO: the equation is not always defined

Number of programs:
\[
  \left( C^{|\predicates{}|} \prod_{\mathsf{P}/a \in \predicates{}} H_a \right)
  \left( \sum_{n=0}^{\maxNumClauses{} - |\predicates{}|} \left( C
    \sum_{a \in \arities{}} p_aH_a \right)^n \right)
\]
The first half of the equation represents setting up a single clause for every
predicate, i.e., we select its body and its head. The second half iterates over
the number of clauses that could be added. For each of the $n$ additional
clauses, we choose its body and its head (defined by a predicate and its arity).

Parameters that affect this:
\begin{itemize}
\item $\arities{}$ and $(p_a)_{a \in \arities{}}$. Consider even zero arity.
\item $|\predicates{}| = \sum_{a \in \arities{}} p_a$
\item $|\variables{}|$
\item $|\constants{}|$. Consider no variables and no constants.
\item $\maxNumNodes{} \ge 1$
\item $\maxNumClauses{} \ge |\predicates{}|$
\end{itemize}
% TODO: this is without any kind of cycle detection and without probabilities
